---
title: "Phylogenetic reconstruction"
author: Walter Xie and Alexei Drummond
output: md_document
---

## Preparation

In this lab, we will learn how to use R to construct phylogenetic trees using some basic algorithms,
such as UPGMA and Neighbour-joining, etc. 

Please download the nucleotide [alignment](ABCD.fasta), and save it into your working directory.
You can preview the alignment by using any text editor to open the file. 

We also need three R libraries: "ape", "phangorn", and "phytools". 
If your R has not got them, please use `install.packages` to install them.

## 1. Pairwise distances

Let's load "ape" first into R:

```{r}
library(ape)
```

Make sure the the alignment file is in your working directory, then load the alignment from the fasta file:

```{r}
# setwd("~/WorkSpace/BioSci700")
alg = read.FASTA("ABCD.fasta")
```

This alignment has 4 taxa and 20 sites. 
The nucleotides were composited to create the distance matrix in [Lecture 5](https://alexeidrummond.org/bayesian_phylo_lectures/lecture5/#/19).

Using the function `dist.dna` to compute the distribution of pairwise distances, 
we can multiple it with the number of sites (20) to obtain the matrix of pairwise distances.


```{r}
# 20 sites
d = dist.dna(alg, model = "raw") * 20
d
```

__Question 1 :__ what does this matrix tell us?

## 2. Algorithms

Secondly, we will use the library "phangorn" to create trees, and use "phytools". 

```{r}
library(phangorn)
library(phytools)
```

### 2.1 UPGMA

Now, please reuse the distance matrix `d` and compute the UPGMA tree. 
Use `plot` function to draw the tree, and use `edgelabels` to show branch lengths.

```{r}
treeUPGMA <- upgma(d)
plot(treeUPGMA, use.edge.length=T, no.margin=TRUE)
edgelabels(treeUPGMA$edge.length)
```


### 2.2 Neighbour-joining

Then compute the neighbour-joining tree. As it produces unrooted trees, 
you need to add the "unrooted" argurment to the `plot` function.

```{r}
treeNJ <- NJ(d)
plot(treeNJ, type="unrooted", use.edge.length=T, no.margin=TRUE)
edgelabels(treeNJ$edge.length)
```

__Question 2 :__ what is pros and cons of the neighbour-joining algorithm?


### 2.3 Parsimony

To understand the parsimony method, we need a more complex example.
Please follow the instructions below, which are modified from the "phangorn" [tutorial](https://cran.r-project.org/web/packages/phangorn/vignettes/Trees.html).

First load the alignment "primates", and remove the outgroup sequence "Bovine" to make the analysis simple. 

```{r}
fdir <- system.file("extdata/trees", package = "phangorn")
primates <- read.phyDat(file.path(fdir, "primates.dna"), format = "interleaved")
names(primates)
# rm "Bovine"
primates <- primates[-2]
names(primates)
```

We can now create the UPGMA and NJ tree respectively.

```{r}
dm  <- dist.ml(primates)
treeUPGMA  <- upgma(dm)
treeNJ  <- NJ(dm)
```

Compare the parsimony score between two trees.

```{r}
parsimony(c(treeUPGMA, treeNJ), primates)
```

We then can give a starting tree, and perform nearest-neighbor interchanges (NNI)
to find the maximum parsimony tree. 

```{r, results='hide'}
treePars <- optim.parsimony(treeUPGMA, primates, rearrangements = "NNI")
```

In the end, plot the tree.

```{r}
plot(treePars, type="unrooted", no.margin=TRUE)
```

__Question 3 :__ assuming there are two trees giving different parsimony scores, which tree should be the maximum parsimony tree? Why?


### 2.4 Maximum likelihood


```{r}
dm <- dist.ml(primates, "F81")
# NJ starting tree
tree <- NJ(dm)
```

```{r, results='hide'}
# HKY + Γ(4)
fitStart <- pml(tree, primates, k=4)
fitHKY <- optim.pml(fitStart, model="HKY", optGamma=TRUE, rearrangement="stochastic")
bs <- bootstrap.pml(fitHKY, bs=100, optNni=TRUE, multicore=TRUE)

plotBS(midpoint(fitHKY$tree), bs, type="phylogram")
```
