---
title: "Step by step"
output:
  html_document:
    code_folding: hide
---


## Function: `get_homology_statements()`

This function extracts the set of **homology statements** implied by a multiple
sequence alignment. Each statement encodes that a specific residue position in
one sequence is evolutionarily equivalent to a specific position in another,
because they co-occur in the same alignment column.

---

## Setup

Before running the exercises, ensure the following packages are loaded and the
helper alignment object is ready. This small two-sequence alignment is used
throughout the worked examples to illustrate how `get_homology_statements()`
maps alignment columns to ungapped residue positions.
```{r setup, class.source="fold-show", message=FALSE, warning=FALSE}
library(msa)
library(Biostrings)

# Two sequences × 4 columns
#
# Col:  1    2    3    4
# S1:   A    C    -    E     <- gap at col 3
# S2:   A    -    D    E     <- gap at col 2
#
# Ungapped position mapping:
#   S1: col 1 → pos 1 (A),  col 2 → pos 2 (C),  col 3 → NA,     col 4 → pos 3 (E)
#   S2: col 1 → pos 1 (A),  col 2 → NA,          col 3 → pos 2 (D), col 4 → pos 3 (E)
#
# Homology statements emitted:
#   col 1: both present → "S1|1|S2|1"
#   col 2: S2 is gap    → skip
#   col 3: S1 is gap    → skip
#   col 4: both present → "S1|3|S2|3"

aln <- as(
  AAMultipleAlignment(c(S1 = "AC-E", S2 = "A-DE")),
  "MsaAAMultipleAlignment"
)
```


### Step 1: Sort sequences by name
  
  Taxa order can differ between alignments produced by different algorithms.
Sorting alphabetically by name guarantees that the same sequence pair always
generates the same statement string, regardless of which aligner was used.

Tips: use the functions `order` and `names`.

```{r step1-sort}
# Convert alignment to AAStringSet and sort sequences alphabetically by name
sorted_set <- as(aln, "AAStringSet")
sorted_set <- sorted_set[order(names(sorted_set))]

# Inspect the sorted sequence names
sorted_set
```

**Q1.** Why would leaving sequences in their original (unsorted) order cause
two alignments of the same sequences to produce non-comparable homology
statement sets?
  
---
  
### Step 2: Convert to a character matrix
  
  A matrix representation gives direct access to any residue by row (sequence)
and column (alignment position), which is required for the column-by-column
scan in later steps.

```{r step2-matrix}
# Convert sorted alignment to a plain character matrix
aln_matrix <- as.matrix(sorted_set)
n_seq      <- nrow(aln_matrix)
n_col      <- ncol(aln_matrix)
seq_names  <- rownames(aln_matrix)

cat("Sequences :", n_seq, "\n")
cat("Columns   :", n_col, "\n")
cat("Names     :", paste(seq_names, collapse = ", "), "\n")
```

**Q2.** What does each row and each column of `aln_matrix` represent?
  What character is used to represent a gap?
  
---
  
### Step 3: Map alignment columns to ungapped positions
  
  For each sequence, build a lookup vector that translates an alignment column
index into the corresponding ungapped residue position. Gap characters (`-`)
are assigned `NA` because they do not represent a real residue.

Tips: use the functions `lapply` and `names`.

```{r step3-ungapped}
# For each sequence, map column index → ungapped residue position (NA for gaps)
col_to_ungapped <- lapply(seq_len(n_seq), function(i) {
  residue_positions <- rep(NA_integer_, n_col)
  ungapped_pos      <- 0L
  for (col in seq_len(n_col)) {
    if (aln_matrix[i, col] != "-") {
      ungapped_pos <- ungapped_pos + 1L
      residue_positions[col] <- ungapped_pos
    }
  }
  residue_positions
})
names(col_to_ungapped) <- seq_names

# Inspect the mapping for the first sequence (first 20 columns)
cat("Column-to-ungapped mapping for", seq_names[1], "(first 20 cols):\n")
print(col_to_ungapped[[seq_names[1]]][1:20])
```

**Q3.** For a sequence `"AC-DE"`, what would `col_to_ungapped` contain
at column 3? At column 4? Explain why.

---
  
### Step 4: Generate all sequence pairs
  
  For *n* sequences there are *n(n−1)/2* unique pairs. Each pair will be
inspected at every alignment column in the next step. 
Apply the function `combn` to find all combinations of the elements.

```{r step4-pairs}
# Enumerate all unique sequence pairs
seq_pairs <- combn(seq_names, 2, simplify = FALSE)

cat("Number of sequence pairs:", length(seq_pairs), "\n")
cat("Expected (n*(n-1)/2)    :", n_seq * (n_seq - 1) / 2, "\n")
```

**Q4.** For 4 sequences, how many unique pairs are there?
  Verify your answer matches the output above.

---
  
### Step 5: Emit homology statements column by column
  
  For every alignment column and every sequence pair, a homology statement is
emitted only when **both** sequences have a real residue at that column.
If either has a gap, no homology is claimed and the pair is skipped.
Each statement is stored as `"seqA|posA|seqB|posB"`.

```{r step5-emit}
# Emit a homology statement for each column where both sequences have a residue
statements <- character(0)

for (col in seq_len(n_col)) {
  for (pair in seq_pairs) {
    seqA <- pair[1]
    seqB <- pair[2]
    posA <- col_to_ungapped[[seqA]][col]
    posB <- col_to_ungapped[[seqB]][col]
    
    if (!is.na(posA) && !is.na(posB)) {
      stmt <- paste(seqA, posA, seqB, posB, sep = "|")
      statements <- c(statements, stmt)
    }
  }
}

cat("Raw statements generated:", length(statements), "\n")
cat("First 6 statements:\n")
print(head(statements, 6))
```

**Q5.** What does the statement `"SeqA|3|SeqB|5"` mean biologically?
  Why is a statement **not** emitted when one sequence has a gap at a column?
  
---
  
### Step 6: Return the unique statement set
  
  Identical alignment columns would produce duplicate statements for the same
position pair. Applying `unique()` ensures the result is a true set, which
is required for the set-theoretic operations in `homology_sym_diff()`.

```{r step6-unique}
# Remove duplicates to obtain the final homology statement set
statements_unique <- unique(statements)

cat("Statements before deduplication:", length(statements),        "\n")
cat("Statements after  deduplication:", length(statements_unique), "\n")
cat("Duplicates removed              :", 
    length(statements) - length(statements_unique), "\n")
```

**Q6.** Under what alignment condition would duplicates appear?
  Why does `homology_sym_diff()` require the input to be a unique set?
  
---


```{r test}
# Homology statements emitted:
#   col 1: both present → "S1|1|S2|1"
#   col 2: S2 is gap    → skip
#   col 3: S1 is gap    → skip
#   col 4: both present → "S1|3|S2|3"

print(statements_unique)
```




