---
title: "BioSci700 Assignment4: Trait Evolution and Phylogenetic Comparative Methods"
author: "Student Name (UPI)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_folding: show
    highlight: tango       # colour the R code
---

<style>
table {
  border-collapse: collapse;
  margin: 1.8em 0;
  width: auto;
}

table th,
table td {
  border: 1px solid #bbb;
  padding: 8px 16px;
  text-align: left;
}

table thead th {
  background-color: #f0f0f0;
  font-weight: bold;
}
</style>

# Assignment4: Trait Evolution and Phylogenetic Comparative Methods (25 points)

**Due date:** Mon 5/6/2026 at 23:59

## Lab Introduction and Objectives

This lab covers phylogenetic comparative methods, building on Chapter 8 of the textbook [*Decoding Genomes*](https://decodinggenomes.org). The central question is how phenotypic traits evolve across species and why shared evolutionary history must be accounted for when drawing biological conclusions.

The four sections progress from testing discrete-trait associations with Fisher's exact test (first assuming independence, then accounting for phylogeny), to applying phylogenetically independent contrasts (PICs) for continuous traits, to fitting and comparing models of continuous trait evolution with `geiger`, and finally to modelling discrete character evolution with the Mk model and stochastic character mapping in `phytools`.

Please follow the submission instructions below before starting:

1. **File naming:** Save this file with your name and student ID: `Lab4_YourName_UPI.Rmd`
2. **Code blocks:** Complete the code in the designated areas marked with `### Your code here` and `### Your code finished above`
3. **Questions:** Answer all questions in the provided text blocks marked with
```
Your answer here
```
4. **Reproducibility:** Ensure all code runs in R 4.2.x or later without errors before submission
5. **Documentation:** Comment your code clearly so that your reasoning is visible to the marker


## Learning Objectives

By the end of this lab, you will be able to:

1. Construct contingency tables and apply Fisher's exact test to assess associations between discrete traits
2. Explain why ignoring phylogenetic relationships can produce spurious conclusions
3. Apply phylogenetically independent contrasts (PICs) to correct for shared ancestry in regression
4. Fit and compare Brownian Motion, Ornstein-Uhlenbeck, Early Burst, and Pagel's lambda models of continuous trait evolution using AIC
5. Fit the Mk model of discrete trait evolution and apply stochastic character mapping to reconstruct character histories with full uncertainty quantification


## Marking

This assignment is worth **25 points** in total:

| Component | Points |
|---|---|
| Section 1: Contingency Table and Fisher's Exact Test | 3 |
| Section 2: Phylogenetically Independent Contrasts | 8 |
| Section 3: Fitting Models of Continuous Trait Evolution | 8 |
| Section 4: Discrete Character Evolution — Mk Model and Stochastic Character Mapping | 4 |
| **Code quality and written answers** | **2** |

**Code quality and written answers (2 points):** One point is awarded for clean, readable code, and the second point is awarded for the quality of your written answers.


---


## Lab Preparation

Complete the following **before** your lab session.

### Required reading

Read **Chapter 8 — Traits and Comparative Methods** of the course textbook [*Decoding Genomes*](https://decodinggenomes.org). Understand the following concepts:

- null hypothesis H₀
- contingency table
- Fisher's test
- odds ratio, and true odds ratio


### Install required R packages

Install the following R packages: "ape", "geiger", "phytools", "tidyverse".


---


# Set up

We will use the following R libraries in this lab.

```{r setup, message=FALSE, warning=FALSE}
library(ape)
library(geiger)
library(phytools)
library(tidyverse)

set.seed(700)
# please replace to your working dir
knitr::opts_knit$set(root.dir = "~/WorkSpace/BioSci700/Lab4")
```


---


# Section 1: Contingency Table and Fisher's Exact Test (3 points)

## 1.1 Reproduce the textbook bird example

### Background

When studying whether two discrete traits are associated across species, a natural starting point is to count how often each combination of trait states occurs and to test whether the pattern deviates from what we would expect by chance. A **contingency table** summarises these counts, and **Fisher's exact test** computes the exact probability of observing the data under a null hypothesis H₀. For this example, the textbook states H₀ as: *having yellow eyes is equally likely among green- and blue-feathered species.*

The [textbook](https://decodinggenomes.org) (Section 8.1.1) illustrates this with eight bird species whose eye colour (yellow or red) and feather colour (green or blue) are recorded. Four species have yellow eyes and green feathers; four have red eyes and blue feathers, a perfectly correlated pattern. The textbook computes the p-value using the hypergeometric distribution, where Z is the number of species with both yellow eyes and green feathers (Equation 8.1):

$$P(Z = 0 \mid H_0) = \frac{\binom{4}{4}\binom{4}{0}}{\binom{8}{4}} \approx 0.0143$$

Here Z = 0 means no species have yellow eyes and *blue* feathers. That is, all four yellow-eyed species have green feathers, which is the observed and most extreme possible outcome. We will reproduce this analysis using R below.

The eight species and their discrete characters are:

| Species | Eye | Feather |
|---------|-----------|----------------|
| 1 | yellow | green |
| 2 | yellow | green |
| 3 | yellow | green |
| 4 | yellow | green |
| 5 | red | blue |
| 6 | red | blue |
| 7 | red | blue |
| 8 | red | blue |

Build the contingency table using R:

```{r s1-contingency-extreme}
# Textbook example: perfectly correlated discrete traits
eye     <- c("yellow", "yellow", "yellow", "yellow", "red", "red", "red", "red")
feather <- c("green",  "green",  "green",  "green",  "blue","blue","blue","blue")

# Build the contingency table
ct_extreme <- table(eye, feather)
print(ct_extreme)
```

**Q1 (2 points):** Run Fisher's exact test on `ct_extreme` twice using `fisher.test()`:

- once with the default `alternative = "two.sided"`,
- once with `alternative = "greater"`, which counts only outcomes with *at least as many* yellow+green co-occurrences as observed (i.e., more co-occurrence than expected under H₀).

Print both p-values clearly labelled.

```{r q1-fisher-extreme}
### Your code here

### Your code finished above
```

In your written answer:

1. State whether H₀ is rejected at α = 0.05 for each test.
2. The textbook (Equation 8.1) reports p ≈ 0.0143. Identify which of your two p-values matches this, and explain why the two p-values differ in a couple of sentences.

```
Your answer here
```


### Understanding the `fisher.test()` output

`fisher.test()` reports: *"alternative hypothesis: true odds ratio is not equal to 1"*. Two terms in the output need definition.

**Odds ratio.** The odds ratio (OR) measures the strength of association between two binary variables. An OR of 1 means the traits are independent; OR > 1 means positive association; OR < 1 means negative association. In a 2×2 table, the sample OR is the cross-product of the diagonal cells: (n₁₁ × n₂₂) / (n₁₂ × n₂₁).

**"True" odds ratio.** The word *true* distinguishes the **population parameter** from the **sample statistic**. The sample OR is computed from the observed data; the true OR is the underlying association that exists in the population from which the data were drawn. Fisher's exact test uses the sample data to ask: *if the true population OR were exactly 1 (complete independence), how probable is it to observe a table as extreme as this one?* That probability is the p-value. The test is performed on the sample, but the hypothesis is about the population.

**Anatomy of the R output.**

1. **The Hypothesis (Population/True):** `alternative hypothesis: true odds ratio is not equal to 1`, stating H₀ (OR = 1, no association in the population) and H₁ (OR ≠ 1).
2. **The Confidence Interval (Population/True):** `95 percent confidence interval`, a range of plausible values for the true population OR consistent with the data.
3. **The Sample Estimate (Sample):** `sample estimates: odds ratio`, the OR calculated directly from the observed cell counts; an estimate of the true OR, not the quantity being tested.


## 1.2 Phylogenetic independence and its consequences

### Background

The analysis in Section 1.1 treats each species as an independent observation. But related species share a common ancestor and therefore inherit similar traits, and they are **not** independent. The apparent correlation between eye and feather colour may simply reflect a single shared evolutionary event rather than a genuine biological association.

The [textbook](https://decodinggenomes.org) (Section 8.1.2) addresses this by shifting the unit of analysis from **species tips** to **branches**: instead of asking how many species have yellow eyes and green feathers, we ask how many branches show a co-occurring change in both traits. This gives a new contingency table (Table 8.2):

| | Eye colour changed | Eye colour unchanged | Totals |
|:------------------------|---|---|---|
| Feather colour changed | 1 | 0 | 1 |
| Feather colour unchanged | 0 | 13 | 13 |
| **Totals** | 1 | 13 | 14 |

The new H₀ is: *the number of branches with a change in both feather and eye colour follows a hypergeometric distribution* (i.e., the two changes are equally likely to occur on any branch, independently). Applying Fisher's exact test to this branch-based table (Equation 8.2):

$$P(\text{both traits change on the same branch} \mid H_0) = \frac{\binom{1}{1}\binom{13}{0}}{\binom{14}{1}} \approx 0.0714$$

Since p = 0.0714 > 0.05, we **fail to reject** H₀, a completely different conclusion from Section 1.1.

Reproduce the branch-based result in R:

```{r s1-branch-test}
# Reproduce Table 8.2: counts of co-occurring trait changes per branch
ct_branch <- matrix(
  c(1, 0, 0, 13),
  nrow = 2,
  dimnames = list(
    feather_changed = c("yes", "no"),
    eye_changed     = c("yes", "no")
  )
)
print(ct_branch)

# One-sided test matching Equation 8.2 (p ≈ 0.0714)
fisher.test(ct_branch, alternative = "greater")
```

**Q2 (1 point):** In 2–3 sentences, explain why treating species as independent observations violates the assumptions of Fisher's exact test, and what the change in p-value from Equation 8.1 (p ≈ 0.0143) to Equation 8.2 (p ≈ 0.0714) reveals about the danger of ignoring phylogenetic relationships.

```
Your answer here
```


---


# Section 2: Phylogenetically Independent Contrasts (8 points)

## Background

The non-independence of species data and the need for phylogenetic correction were formalised by Felsenstein (1985), who introduced phylogenetically independent contrasts, and elaborated by Harvey & Pagel (1991) in *The Comparative Method in Evolutionary Biology*.

### Why cross-species regression fails for species data

When two continuous traits appear correlated across species, the association may be partly or entirely explained by **shared evolutionary history** rather than a genuine biological relationship. A standard linear model assumes all observations are independent, but closely related species inherit similar trait values from a common ancestor, and they are not independent data points.

Consider a simple example: the Geospiza tree has a deep split between large ground finches (deep beak, long culmen, long wing) and small tree/warbler finches (shallow beak, short culmen, short wing). A cross-species regression of beak depth on culmen length sees two tight clusters and returns a strong positive slope with a small p-value. But this "correlation" is largely driven by a single deep clade split, one evolutionary event echoed across all 13 tips. The effective sample size is far smaller than 13, so standard errors are underestimated and p-values are too optimistic.

### How PICs fix the problem

**Phylogenetically independent contrasts (PICs)**, introduced by Felsenstein (1985), transform tip values into **differences at internal nodes** that are statistically independent of each other under **Brownian Motion** (BM) evolution, a model in which trait variance accumulates in proportion to elapsed time, like an undirected random walk with no preferred direction.

At each internal node *k* with daughter lineages *i* and *j*, the standardised contrast is:

$$C_k = \frac{x_i - x_j}{\sqrt{v_i + v_j}}$$

where:

- $x_i$, $x_j$: the trait values (e.g. beak depth) of daughter lineages *i* and *j*
- $v_i$, $v_j$: the branch lengths leading to *i* and *j*. Longer branches allow more evolutionary change, so a larger raw difference is expected and needs to be down-weighted
- $C_k$: the contrast at node *k*, the trait difference between the two daughters, scaled by the expected evolutionary change given their branch lengths

With *n* tip species the tree has *n* − 1 internal nodes, yielding *n* − 1 independent contrasts.

Instead of asking "do species with long culmens also have deep beaks?", PICs ask "when a lineage evolved a longer culmen than its sister lineage, did it also evolve a deeper beak?", repeated at every node. This reframing removes the shared-ancestry problem entirely.

### Why regression through the origin is required

Under Brownian Motion, contrasts have **expected value zero**, with no directional prediction for which daughter evolves higher. Fitting a regression with an intercept would allow the model to absorb a spurious mean-shift in the contrasts, wasting a degree of freedom and violating the BM assumption. The `- 1` in `lm(pic_y ~ pic_x - 1)` enforces the biologically correct constraint.


## 2.1 Load and explore the Geospiza data

We will use the classic **Geospiza** (Darwin's finches) dataset included in `geiger`, which contains five morphological measurements for 13 Galápagos finch species on a dated phylogeny: wing length (`wingL`), tarsus length (`tarsusL`), culmen length (`culmenL`), beak width (`beakW`), and beak depth (`beakD`).

```{r s2-load-data}
data(geospiza)

# Match tree tips to data rows (drops 'olivacea' which lacks trait data)
geo <- treedata(geospiza$phy, geospiza$dat, warnings = FALSE)
tree <- geo$phy
traits <- as.data.frame(geo$data)

cat("Tips in tree:", length(tree$tip.label), "\n")
cat("Traits available:", colnames(traits), "\n")
```

**Q3 (1 point):** Plot the phylogenetic tree with tip labels and print a summary table of the five traits (use `summary()`). 

```{r q3-explore}
### Your code here

### Your code finished above
```

Which species has the largest and smallest `beakD`, and what are the values? Which has the largest and smallest `culmenL`, and what are the values? Does the same species tend to sit at both extremes?

```
Your answer here
```


## 2.2 Cross-species regression (ignoring phylogeny)

We will test whether culmen length (`culmenL`) predicts beak depth (`beakD`) across finch species.

**Q4 (2 points):** Fit a simple linear regression `beakD ~ culmenL` using `lm()`. Plot the data points with species labels, overlay the regression line, and report **R²**, the **slope**, and its **p-value**. 

*Hint:* Use `data = traits` inside `lm`. Print a summary of regression. For species labels on the plot, use `text()` or `geom_text()` if using ggplot2.

```{r q4-cross-species-regression}
### Your code here

### Your code finished above
```

Does culmen length appear to predict beak depth significantly? Do any species appear to be influential points, and how might they affect the slope estimate?

```
Your answer here
```


## 2.3 Phylogenetically independent contrasts

**Q5 (2 points):** Compute PICs for both `culmenL` and `beakD` using `ape::pic()`, then regress the contrasts through the **origin**, that is, fit a no-intercept model (`lm(pic_beakD ~ pic_culmen - 1)`), because under Brownian Motion the expected contrast is zero and allowing an intercept would violate that assumption. Print a summary of the regression and plot the result with dashed reference lines using `abline(h = 0, v = 0, lty = 2)`, which mark nodes where neither trait changed.

*Note:* `summary()` will report an R² value. For a no-intercept model (fitted with `- 1`), R internally computes R² as `1 − SS_res / sum(y²)` rather than the usual `1 − SS_res / SS_centered`, which inflates the value, so this R² is not comparable to the one in Q4. You will explain this in Q6.

*Hint:* `pic()` requires a named numeric vector (provided as scaffolding) and returns *n* − 1 branch-length standardised contrasts, one per internal node. Call `pic()` on each named vector, then fit the no-intercept model described above.

```{r q5-pic-regression}
culmen_vec <- setNames(traits$culmenL, rownames(traits))
beakD_vec  <- setNames(traits$beakD,   rownames(traits))

### Your code here

### Your code finished above
```

Report the slope and p-value of the PIC regression, and explain biologically what the result tells you about the relationship between culmen length and beak depth after accounting for shared evolutionary history.

```
Your answer here
```


## 2.4 Compare cross-species vs PIC results

**Q6 (3 points):** Extend your analysis to a second trait pair. Run both the cross-species `lm()` and PIC regression for `beakD ~ wingL` and fill in the comparison table below.

*Note on R²:* Do **not** include R² in your comparison. The PIC regression is fitted through the origin, so R's `summary()` computes R² relative to zero rather than relative to the mean, making it numerically incomparable to the cross-species R². Use slope and p-value instead.

*Hint:* `wingL` is a column in `traits`. The PIC workflow is identical to Q5. Reuse `beakD_vec` and `pic_beakD`, and add a `wingL` vector.


- **(1 point)** Implement R code, and correctly compute and fill in slope and p-value for all four rows.

```{r q6-second-pair}
### Your code here

### Your code finished above
```

Fill in the table with your results (round to 3 decimal places):

| Trait pair | Method | Slope | p-value |
|---|---|---|---|
| `beakD ~ culmenL` | Cross-species `lm` | | |
| `beakD ~ culmenL` | PIC `lm` | | |
| `beakD ~ wingL` | Cross-species `lm` | | |
| `beakD ~ wingL` | PIC `lm` | | |

- **(1 point)** Are both trait pairs still significant under PIC? Does significance increase, decrease, or stay the same? Give a biological reason for the pattern you observe. (*Hint:* consider whether each trait pair shares a common functional or size-based explanation for their co-evolution.)
- **(1 point)** Explain in 2–3 sentences why PIC regression is statistically preferred over cross-species regression even when both give significant results, using the concept of non-independence of species data.

```
Your answer here
```


---


# Section 3: Fitting Models of Continuous Trait Evolution (8 points)

## Background

Different traits do not evolve the same way, and assuming a single model without testing alternatives can lead to wrong conclusions about adaptation, selection, and evolutionary rates. Maximum likelihood fitting lets us evaluate several competing models on the same data and identify which the data best support, penalised for complexity. The `geiger` package implements this for four standard models:

| Model | Description | Key parameter |
|-------|-------------|---------------|
| **BM** | Brownian Motion, in which trait drifts randomly with no directional force | σ² (rate) |
| **OU** | Ornstein-Uhlenbeck, where trait is pulled towards an optimum θ by strength α | α (pull strength), σ², θ |
| **EB** | Early Burst, where rate decelerates exponentially over time | `a` (rate-change parameter; `a` < 0 = deceleration) |
| **lambda** | Pagel's λ, which scales internal branch lengths to test phylogenetic signal | λ (0 = no signal, 1 = pure BM) |

We use AIC to compare non-nested models: the model with the lowest AIC is preferred, and a difference ΔAIC > 2 indicates meaningful support for the better model (Section 7.3 of the [textbook](https://decodinggenomes.org)).


## 3.1 Fit the four models to beak depth

**Q7 (3 points):** Using `geiger::fitContinuous()`, fit all four models (BM, OU, EB, lambda) to the `beakD` trait from the Geospiza dataset. For example `fit_bm <- fitContinuous(tree, beakD_vec, model = "BM")`. For each model, print the log-likelihood, number of parameters (`k`), and AIC. (Use `fit$opt$lnL`, `fit$opt$k`, and `fit$opt$aic`.)
*Note:* The OU model may print a warning `Parameter estimates appear at bounds: alpha`. This means the optimiser could not find an interior maximum, so the OU fit is effectively degenerating to BM. Report it in your table and discuss it in Q9.

```{r q7-fit-models}
beakD_vec <- setNames(traits$beakD, rownames(traits))  # repeated from Section 2 for standalone use

### Your code here

### Your code finished above

```

In your written answer, briefly state what `k` represents and why it matters for the AIC calculation.

```
Your answer here (table of lnL, k, AIC for each model)
```


## 3.2 AIC-based model comparison

**Q8 (3 points):** Collect the four AIC values into a named vector and pass them to `geiger::aicw()` to obtain AIC weights. For example,

```r
aic_vals <- c(BM = fit_bm$opt$aic, OU = fit_ou$opt$aic, EB = fit_eb$opt$aic, lambda = fit_lambda$opt$aic)
aicw(aic_vals)
```

In the end, display the full comparison table (AIC, ΔAIC, weight). 

```{r q8-AIC-model-comparison}
### Your code here

### Your code finished above
```

Which model has the strongest support? Is the winning model's AIC weight > 0.9, indicating clear support, or is evidence spread across models? (As a guide: weight > 0.95 indicates near-decisive support; no model exceeding 0.7 indicates substantial uncertainty across models.)

```
Your answer here
```


## 3.3 Interpret the best model's parameters

**Q9 (2 points):** For the best-supported model identified in Q8, extract and report all estimated parameters.
For example, `print(fit_bm$opt)` if you think BM is the best-supported model.

*Parameter names by model:* BM → `fit$opt$sigsq`; OU → `fit$opt$alpha`, `fit$opt$sigsq`; EB → `fit$opt$a`, `fit$opt$sigsq`; lambda → `fit$opt$lambda`, `fit$opt$sigsq`. All models also return `fit$opt$z0` (root state estimate).

```{r q9-interpret}
### Your code here
# Print the opt list of the best model:

### Your code finished above
```

- **(1 point)** Identify the key diagnostic parameter of the winning model (λ for lambda, α for OU, `a` for EB, σ² for BM). Explain what its biological extremes mean, and state where your estimated value falls on that spectrum.
- **(1 point)** Give a biological interpretation: what does the winning model tell us about how beak depth evolved in Darwin's finches, and does this make sense given the ecological diversity of this group?

```
Your answer here
```


---


# Section 4: Discrete Character Evolution — The Mk Model and Stochastic Character Mapping (4 points)

## Background

### The Mk model

Discrete traits evolve through transitions between states rather than continuous drift. The **Mk model** (Markov k-state model; Lewis 2001) is the standard framework: it treats transitions among k discrete states as a continuous-time Markov process governed by an instantaneous rate matrix **Q**, where each off-diagonal element q_ij is the rate of transitioning from state i to state j per unit branch length. For a binary trait (deep / shallow), the Q matrix takes the form:

$$Q = \begin{pmatrix} -q_{ds} & q_{ds} \\ q_{sd} & -q_{sd} \end{pmatrix}$$

Two common special cases are **ER (Equal Rates)**, where q_ds = q_sd = q (one free parameter), and **ARD (All Rates Different)**, where q_ds and q_sd are estimated separately (two free parameters). `phytools::fitMk()` fits the Mk model by maximum likelihood and returns the full estimated Q matrix, making it possible to inspect not only which model fits better but how fast and in which direction transitions occur.

### Stochastic character mapping

A limitation of deterministic ancestral state reconstruction (such as `ace()`) is that it returns a single point estimate of the most likely state at each node, hiding the true uncertainty in character history. **Stochastic character mapping** (Huelsenbeck et al. 2003) addresses this by sampling many complete character histories from the posterior, where each history specifies not just node states but the exact timing of every transition along every branch. Averaging across N sampled histories gives node probabilities that properly account for this uncertainty. `phytools::make.simmap()` generates N stochastic maps; `summary()` extracts node probabilities averaged across all of them.


## 4.1 Create a binary trait

We classify each finch species as having a "deep" beak (beakD ≥ median) or a "shallow" beak (beakD < median), representing the broad ecological split between seed-crushing ground finches and insect-probing warbler finches.

```{r s4-binary-trait}
median_beakD <- median(traits$beakD)
beak_type <- ifelse(traits$beakD >= median_beakD, "deep", "shallow")
beak_type <- setNames(beak_type, rownames(traits))

cat("Median beakD:", round(median_beakD, 3), "\n")
print(beak_type)
```


## 4.2 Fit the Mk model

**Q10 (2 points):** Fit the ER and ARD Mk models to `beak_type` using `phytools::fitMk()`. Print the estimated Q matrix for each model using `print()`, and compare their AIC values using `AIC()`.

*Hint:* `fitMk()` takes the same `tree` and named character vector `beak_type` used throughout this lab, and a `model` argument of `"ER"` or `"ARD"`.

```{r q10-fitMk}
### Your code here

### Your code finished above
```

Which model is better supported? Do the forward (deep → shallow) and reverse (shallow → deep) rates differ substantially under ARD? If the estimated rates are very large (>> 1 per unit branch length), discuss what this implies for the reliability of the model fit and the suitability of this binary trait.

```
Your answer here
```


## 4.3 Stochastic character mapping

**Q11 (2 points):** Use `phytools::make.simmap()` with the better-supported model from Q10 to generate 100 stochastic character maps of `beak_type` on the tree (`nsim = 100`, `Q = "empirical"`). Summarise the maps with `summary()` and extract the averaged node probabilities from the `$ace` element of the summary object. Plot one representative map with `phytools::plotSimmap()` and overlay pie charts at internal nodes showing the averaged node probabilities.

*Hint:* Use `set.seed(700)` before `make.simmap()` for reproducibility. `plotSimmap()` takes a `colors` argument (named vector of state colours). Use `title()` separately to add a plot title. Pass `piecol` to `nodelabels()` in the same order as the column names of the summary `$ace` matrix.

```{r q11-simmap, fig.width=8, fig.height=6}
set.seed(700)

### Your code here

### Your code finished above
```

What probability does the stochastic mapping assign to each state at the root, averaged across the 100 simulations? Is the reconstruction confident? Explain why stochastic character mapping is more informative than a single-history reconstruction from `ace()`.

```
Your answer here
```


---


## Submission Checklist

- File is renamed with your name and student ID: `Lab4_YourName_UPI.Rmd`, including either HTML or PDF after Knit.
- All code blocks are completed and run without errors
- All questions are answered with clear and thoughtful responses
- All plots and printed outputs are visible in the knitted document
- The document knits to HTML without errors

---

*Submit both the .Rmd file and the knitted .html or .pdf file through the course management system.*
